Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads

### 2023-05-02 Huston_Paper_asme.tex
formatting + moar stuff

\documentclass[journal,onecolumn]{IEEEtran}

\hyphenation{op-tical net-works semi-conduc-tor}
\usepackage[square,sort&compress,numbers]{natbib}
\usepackage[margin=1in]{geometry}
\usepackage{textcomp} 
\usepackage{amssymb}  
\usepackage{bm}       
\usepackage{booktabs}
\usepackage{dcolumn}  
\usepackage{ragged2e}
\usepackage{graphicx} 
\usepackage{float}
\usepackage{rotating}
\usepackage{makecell}
\usepackage{multirow}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{array}
\usepackage[printonlyused]{acronym}
\usepackage[framed, numbered]{matlab-prettifier}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{times}
\usepackage{indentfirst}
\usepackage{multicol}
\usepackage{enumitem}
\usepackage{attachfile}
%\usepackage{tabularray}

% Table Stuff
\newcommand{\spheading}[2][9.5em]{% \spheading[<width>]{<stuff>}
	\rotatebox{90}{\parbox{#1}{\raggedright #2}}}

\begin{document}
	%
	% paper title
	% Titles are generally capitalized except for words such as a, an, and, as,
	% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
	% not capitalized unless they are the first or last word of the title.
	% Linebreaks \\ can be used within to get better formatting as desired.
	% Do not put math or special symbols in the title.
	\title{Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads}
	%
	%
	% author names and IEEE memberships
	% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
	% a structure at a ~ so this keeps an author's name from being broken across
	% two lines.
	% use \thanks{} to gain access to the first footnote area
	% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
	% was not built to handle multiple paragraphs
	%

	\author{Rhett Huston, Dr. Jay Wilhelm}
%	{
%		and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
%		\thanks{M. Shell was with the Department
%			of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
%			GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
%		\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
%		\thanks{Manuscript received April 19, 2005; revised August 26, 2015.}}

	% note the % following the last \IEEEmembership and also \thanks - 
	% these prevent an unwanted space from occurring between the last author name
	% and the end of the author line. i.e., if you had this:
	% 
	% \author{....lastname \thanks{...} \thanks{...} }
	%                     ^------------^------------^----Do not want these spaces!
	%
	% a space would be appended to the last name and could cause every name on that
	% line to be shifted left slightly. This is one of those "LaTeX things". For
	% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
	% "AB" then you have to do: "\textbf{A}\textbf{B}"
	% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.

% The paper headers
\markboth{Journal of Autonomous Vehicles and Systems}%
{}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

% If you want to put a publisher's ID mark on the page you can do it like
% this:
%\IEEEpubid{0000--0000/00\$00.00~\copyright~2015 IEEE}
% Remember, if you use this you must call \IEEEpubidadjcol in the second
% column for its text to clear the IEEEpubid mark.

% use for special paper notices
%\IEEEspecialpapernotice{(Invited Paper)}

% make the title area
\maketitle

% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\begin{abstract}{

	{Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles. Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough. Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching. This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces. Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads. The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations. Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces. It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively. Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy. Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}	
}	
\end{abstract}

%	
%	% Note that keywords are not normally used for peerreview papers.
%	\begin{IEEEkeywords}
%		IEEE, IEEEtran, journal, \LaTeX, paper, template.
%	\end{IEEEkeywords}
%	
%	
%	
%

@@

	%Real time description of surface noise based on incoming LiDAR data, then subsequent comparison to a standard, is a possible alternative that may allow for computationally efficient road surface detection. Overlaying camera imagery data with LiDAR allows for assigning incoming LiDAR data with color information, potentially increasing efficiency by narrowing the search area.

	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident. This work achieved 91.67\% intercepting gravel road detection accuracy with a 1.226 seconds per 360 degree scan classification rate.}

@@

	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff. LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}). Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using Robotic Operating System (ROS). Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio. Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot. Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot. Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection. Only three classes are considered in this work: "gravel", "asphalt", and "unknown". }

	\begin{figure}[H] 
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
			\caption[Sensor Van]{}
			\label{fig:van}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.9\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
			\caption[VLP 32 on Van]{}
			\label{fig:vlp32mount}
		\end{subfigure}
		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
		\label{fig:Experimental_Apperatus}
	\end{figure}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
			\caption[Blackburn Road Camera View]{}
			\label{fig:Blackburn_Road_View}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/gravel_lot_pic.png}
			\caption[Armig Road Camera View]{}
			\label{fig:Gravel_Lot_View}
		\end{subfigure}
		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
		\label{fig:Combined_Roads}
	\end{figure}

@@

	{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds. Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converts the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}). While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes. Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}). Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.}

	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
			\caption[C]{}
			\label{fig:Compiled_PCD}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
			\caption[]{}
			\label{fig:Classified_Areas_PCD}
		\end{subfigure}
		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a). Manually defined areas inform where training data lies (b).}
		\label{fig:Combined_Classified_PCD}
	\end{figure}

@@

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified. Shown is the results of one run. Percentages of each terrain type per area were calculated. Area $3$ is the gravel parking lot entrance from which training data was extracted.}
		\label{fig:area_percentages}
	\end{figure}

@@

	\begin{figure}[H]
			\centering
			\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
			\caption[Post-Adjusted]{}
			\label{fig:prepostadjust}
		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm efficiency of unmarked road detection. }
	\end{figure}

@@

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_toc}
		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed using a rule set and compared to actual road-surface areas. }
		\label{fig:rm_db_4_toc}
	\end{figure}

@@

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

% use appendices with more than one appendix
% then use \section to start each appendix
% you must declare a \section before using any
% \subsection or using \label (\appendices by itself
% starts a section numbered zero.)
%

\appendices
%\section{Equations}

%\begin{equation}\label{eqn:MLS}
%	\left[ {\begin{array}{cc}
%			\sum_{i=1}^{m} x_i z_i \\
%			\sum_{i=1}^{m} y_i z_i \\
%			\sum_{i=1}^{m} z_i \\
%			
%	\end{array} } \right]
%	=
%	\left[ {\begin{array}{ccc}
%			\sum_{i=1}^{m} x_i^2 		& \sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} x_i \\
%			\sum_{i=1}^{m} x_i y_i 		& \sum_{i=1}^{m} y_i^2 			& \sum_{i=1}^{m} y_i \\
%			\sum_{i=1}^{m} x_i 			& \sum_{i=1}^{m} y_i 			& \sum_{i=1}^{m} 1   \\
%	\end{array} } \right]
%	\left[ {\begin{array}{cc}
%			A\\
%			B\\
%			C\\
%	\end{array} } \right]
%\end{equation}
%\centering
%{Method of Least Squares Planar Projection}

% you can choose not to have a title for an appendix
%% if you want by leaving the argument blank
%\section{}
%Appendix two text goes here.

% use section* for acknowledgment
\justifying
\section*{Acknowledgment}

	{Special thanks to Dr. Jay Wilhelm for his guidance and Travis Moleski for his help in LiDAR map building.}

% Can use something like this to put references on a page
% by themselves when using endfloat and the captionsoff option.
\ifCLASSOPTIONcaptionsoff
\newpage
\fi

% trigger a \newpage just before the given reference
% number - used to balance the columns on the last page
% adjust value as needed - may need to be readjusted if
% the document is modified later
%\IEEEtriggeratref{8}
% The "triggered" command can be changed if desired:
%\IEEEtriggercmd{\enlargethispage{-5in}}

% references section

% can use a bibliography generated by BibTeX as a .bbl file
% BibTeX documentation can be easily obtained at:
% http://mirror.ctan.org/biblio/bibtex/contrib/doc/
% The IEEEtran BibTeX style support page is at:
% http://www.michaelshell.org/tex/ieeetran/bibtex/
%\bibliographystyle{IEEEtran}
% argument is your BibTeX string definitions and bibliography database(s)
%\bibliography{IEEEabrv,../bib/paper}
%
% <OR> manually copy in the resultant .bbl file
% set second argument of \begin to the number of references
% (used to reserve space for the reference number labels box)

	\bibliographystyle{ieeetr}  
	\bibliography{resources.bib}

% that's all folks
\end{document}

>>>>

\documentclass[balance,upint,subscriptcorrection,varvw,mathalfa=cal=boondoxo,colorlinks]{asmeconf}

%\hyphenation{op-tical net-works semi-conduc-tor}
%\usepackage[square,sort&compress,numbers]{natbib}
%\usepackage{amssymb}  
%\usepackage{bm}       
%\usepackage{booktabs}
%\usepackage{dcolumn}  
%\usepackage{ragged2e}
%\usepackage{graphicx} 
%\usepackage{float}
%\usepackage{rotating}
%\usepackage{makecell}
%\usepackage{multirow}
%\usepackage[english]{babel}
%\usepackage{subcaption}
%\usepackage{array}
%\usepackage[printonlyused]{acronym}
%\usepackage[framed, numbered]{matlab-prettifier}
%\usepackage[utf8]{inputenc}
%\usepackage[english]{babel}

%%%%%  pdf metadata  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypersetup{%
pdfauthor={John H. Lienhard},
pdftitle={ASME Conference Paper LaTeX Template},                  % <=== change to YOUR pdf file title
pdfkeywords={ASME conference paper, LaTeX template, BibTeX style},% <=== change to YOUR pdf keywords
pdfsubject = {Describes the asmeconf LaTeX template},			  % <=== change to YOUR subject
%	pdfurl={https://ctan.org/pkg/asmeconf},% may delete
pdflicenseurl={https://ctan.org/pkg/asmeconf},% may delete
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

% Change these fields to the right content for your conference.
% You can comment these out if for some reason you don't want a header.
% Use title case (first letters capitalized), not all capitals

\ConfName{Proceedings of the ASME 2023\linebreak International Mechanical Engineering Congress and Exposition}
\ConfAcronym{IMECE2023}
\ConfDate{October 29--November 2, 2023} % update 
\ConfCity{New Orleans, LA} % update 
\PaperNo{IMECE2023-XXXX}

% Units of measure (e.g., cm) and other specialty lowercase terms in the title should be 
%   enclosed in \NoCaseChange{...} to maintain lower case type
%   LaTeX will automatically set the rest of the title in all capital letters.

\title{Enhance Road Detection Data Processing of LiDAR Point Clouds to Specifically Identify Unmarked Gravel Rural Roads} % <=== replace with YOUR title
%\title{Place Title Here: Place Subtitle After Colon}

%   Put author names into the order you want. Use the same order for affiliations.
%   \affil{#} tags the author's affiliation to the address in \SetAffiliation{#}.
%   No space between last name and \affil{#}, separate names with commas.
%
%	For a sole author or a single affiliation for all authors, {#} may be left empty, as \affil{} and \SetAffiliation{} (but not with [grid] option!)
%
%   \CorrespondingAuthor{email} follows that author's affiliation, no spaces.  
%   If multiple corresponding authors, put both email addresses in the same command and place after both authors.
%
%   \JointFirstAuthor, if applicable, follows the affiliation of the relevant authors, no spaces.

\SetAuthors{
	Rhett Huston\affil{1}\JointFirstAuthor, 
	Dr. Jay Wilhelm\affil{2}\JointFirstAuthor, 
}

\SetAffiliation{1}{Ohio University, Athens, Ohio}
\SetAffiliation{2}{Ohio University, Athens, Ohio}

\maketitle

%%% Use this footnote for tracking various versions of your draft. Change text to suit your own needs. 
%%% \date{..} calls the same command. 
\versionfootnote{Documentation for \texttt{asmeconf.cls}: Version~\versionno, \today.}% <=== Delete before final submission.

%%% Change these to your keywords.  Keywords are automatically printed at the end of the abstract.
%%% This command MUST COME BEFORE the end of the abstract.
%%% If you don't want keywords, leave the argument of \keywords{} empty (or use the abstract* environment)

%	\keywords{ASME conference paper, \LaTeX\ Template, \hologo{BibTeX}}

%%%%%  End of fields to be completed. Now write your paper. %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%  ABSTRACT  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Abstract should be 200 words or less
\begin{abstract}{

		{Gravel roads lack standardized features such as curbs or painted lines, presenting detection challenges to autonomous vehicles. Global Positioning Service (GPS) and high resolution maps may not be reliable for navigation of gravel roads, as some may only be width of the vehicle and GPS may not be accurate enough. Normal Distribution Transform (NDT) based on scanning LiDAR data may be insufficient for navigating on gravel roads as there may not be enough geometrically distinct features for reliable scan matching. This work will examine a method of classifying scanning LiDAR data spatial and remission features for explicit detection of unmarked gravel road surfaces. Exploration of terrain classification using high resolution scanning LiDAR data of these road surfaces may allow for predicting gravel road boundary locations potentially enabling confident autonomous operations on gravel roads. The principal outcome of this work was a method for gravel road detection using LiDAR data for the purpose of predicting road boundary locations. Random Decision Forests were trained using scanning LiDAR data terrain classification to detect unmarked gravel and asphalt surfaces. It was found that a true-positive accuracy for gravel and asphalt surfaces was 67.54\% and 87.05\% respectively. Overlapping results between manually projected and actual road surface areas resulted in 91.67\% intercepting gravel road detection accuracy. Detection of unmarked road surfaces would increase the operational region capabilities of self driving vehicles considerably by allowing autonomous operations on 1.5 million miles of previously undetected roads.}

	}	
\end{abstract}

%%%%%%%%%  NOMENCLATURE (OPTIONAL) %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% To change space between the symbols and  definitions, use \begin{nomenclature}[Xcm] where X is a number 
	%% The unit cm can be replaced by any LaTeX unit of dimension: pt, in, ex, em, pc, etc.
	%% Default is 2em.
	%% \EntryHeading{..} produces an italicized subheading in the nomenclature list, e.g., \EntryHeading{Greek letters}

	%%%%%%%%%  BODY OF PAPER %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

	{Detection of unmarked road surfaces may utilize roughness properties without distinct markings or topographical features. Surface roughness is a measurable property that may be used to characterize a gravel road, as they are typically consistent with gravel type used \cite{skorseth_gravel_nodate} which may be exploited in a search and compare function, and is distinct from chipseal, grass, and other ground types \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain}. Surface roughness properties may be derived from processing LiDAR spatial and remission data from an aerial and surface level perspective \cite{wan_road_2007, levi_3d_2012_light, levi_3d_2012_terrain, pollyea_experimental_2012,rychkov_computational_2012,lague_accurate_2013,brubaker_use_2013,turner_estimation_2014,campbell_lidar-based_2017,shepard_roughness_2001,tegowski_statistical_2016,sock_probabilistic_2016,milenkovic_roughness_2018,yadav_extraction_2017, yadav_rural_2018}. Implementing a terrain classification model using Random Decision Forests (RDF) using scanning LiDAR data to exploit surface roughness was studied for road surface detection in this study. Evaluation of the detection model will rely principally upon accuracy and processing time, as a moving vehicle must obtain reliable information of the road surface in a timely manner in order to inform trajectory updates, or else risk suffering an accident.}

@@

	\subsubsection{Overview}

		{Followed methodology to accomplish this work is outlined in Figure \ref{fig:flowz_2}. The experimental apperatus for this work is the Ohio University Autonomous Van van (Figure \ref{fig:van}), which has a Velodyne VLP-32C scanning LiDAR sensor, a Novatel PwrPak 7D-E2 GNSS and INS system, and Mako cameras. Raw data was gathered using the Robotic Operating System (ROS) to package incoming scanning LiDAR, GNSS, IMU, and camera sensor data as rosbags. Scanning LiDAR data was sent from the sensor as a single 360 degree scan at 20 Hz. GNSS and IMU data was captured at a rate of 100 Hz. Rosbags were unpacked using MATLAB tools to separate data streams. Prior to training data extraction, manual dictation of gravel and asphalt areas is necessary. In order to do so, all scanning LiDAR data from a rosbag was compiled in order to create a single point cloud map. Scanning LiDAR timestamps were matched to the closest GNSS and IMU timestamp data. Closest matching GNSS and IMU data informed the location and orientation of the LiDAR scan (Figure \ref{fig:Combined_Classified_PCD}). Compiled point cloud maps may then be examined and compared to camera and satellite data to determine gravel and asphalt surface location. Manually defined 2D areas representing gravel and asphalt were projected unto the point cloud (Figure \ref{fig:Classified_Areas_PCD}).}

		{A single fixed arc from three scanning LiDAR channels in front of the experimental apparatus were considered for classification. Future work may incorporate additional channels. Training data was extracted by examining each 360 scan and comparing the location of the three arcs and the manually projected 2D areas. If the arc resides within the manually defined surface area, raw spatial and intensity values from the arc were saved to a raw training database. Each channel had a separate training data base. Features that describe the raw spatial and intensity data were extracted by performing mathematical functions and saving the results to a feature data base. Thirty percent of the training data base was randomly selected and set aside as a validation data base. Training data was visually examined for clustering to verify that a Random Decision Forest would be able to be trained (Figure \ref{fig:training_data_cluster}). It was found that each class generally clustered to a certain range with some overlap, which indicates that a Random Decision Forest may be trained. MATLAB was used to train a Random Decision Forests for each of the three considered channels using Bayesian Optimization to tune tree depth and number of splits. Training continued until a minimum error was reached (Figure \ref{fig:c2_min_class_error}).}

		{Validation of the }

		%When the app performs hyperparameter tuning by using Bayesian optimization (see Optimization Options for a brief introduction), it chooses the set ofhyperparameter values that minimizes an upper confidence interval of the classification error objective model, rather than the set that minimizes the classification error.

	\subsection{Detailed Methodology}

	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{figures/flowz_2}
		\caption[Project Flow]{High level view of project work}
		\label{fig:flowz_2}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{figures/training_data_cluster}
		\caption[Example Clustering]{Example of training data comparison between two features.}
		\label{fig:training_data_cluster}
	\end{figure}

	\begin{figure}
		\centering
		\includegraphics[width=\linewidth]{figures/c2_min_class_error}
		\caption[RDF Training Classification Error]{}
		\label{fig:c2_min_class_error}
	\end{figure}

	{Required physical data was gathered by the Ohio University Autonomous Van van (Figure \ref{fig:van}), built by AutonomouStuff. LiDAR point cloud data was gathered with the vehicle's roof-mounted Velodyne VLP-32C (Figure \ref{fig:vlp32mount}). Velodnyne's VLP-32C has 32 channels producing 300,000 points per second with a vertical field of view from -45$^{\circ}$ to $+$15$^{\circ}$, providing a high output data stream \cite{vlp_32c}. GPS data was recorded with a Novatel PwrPak 7D-E2, a Global Navigation Satellite System (GNSS) \& Inertial Navigation System (INS), and two GNSS-502 antennas by NavtechGPS. Data acquisition and recording was completed on the vehicle's integrated Ubuntu 18.04 system using ROS. Scanning LiDAR data of a gravel surface was gathered from a gravel parking lot that intersects Blackburn Road near Athens Ohio. Scanning LiDAR data of an asphalt surface was gathered from Blackburn Road near the intersecting gravel lot. Scanning LiDAR data of grassy surfaces was gathered from the side of the gravel lot. Although this work does not evaluate the detection accuracy of grass, differentiation between a road and non-road surface is necessary for evaluating road-surface detection. Only three classes are considered in this work: "gravel", "asphalt", and "unknown". }

	\begin{figure} 
		\centering
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/van_on_van}
			\caption[Sensor Van]{}
			\label{fig:van}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth,height=4.60 cm,keepaspectratio]{figures/LiDAR_on_van}
			\caption[VLP 32 on Van]{}
			\label{fig:vlp32mount}
		\end{subfigure}
		\caption[Experimental Apparatus]{Experimental Apparatus (a). Velodyne VLP-32 Scanning LiDAR is mounted on top of the vehicle (b).}
		\label{fig:Experimental_Apperatus}
	\end{figure}

	\begin{figure}
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/vlcsnap-2023-04-20-08h48m17s447}
			\caption[Blackburn Road Camera View]{}
			\label{fig:Blackburn_Road_View}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\linewidth,height=5.0 cm,keepaspectratio]{figures/gravel_lot_pic.png}
			\caption[Armig Road Camera View]{}
			\label{fig:Gravel_Lot_View}
		\end{subfigure}
		\caption[Blackburn Road \& Gravel Parking Lot]{Road view of Blackburn Road - an unmarked asphalt road (a). Road view of the gravel driveway (b).}
		\label{fig:Combined_Roads}
	\end{figure}

@@

	{Scanning LiDAR training data was extracted from manually defined gravel, asphalt, and unknown terrain areas on compiled point clouds. Compiling LiDAR data into a single point cloud was completed by post processing rosbags that contained scanning LiDAR, GPS, and IMU data of an unmarked road. Point cloud translation and rotation was accomplished using transformation matrices derived from GPS and IMU data. Rotation matrices were created using extracted roll, pitch, and yaw data from the IMU. LiDAR origin was found using GPS longitude, latitude, and altitude data. Physical distances between the GPS, IMU, and LiDAR was rectitude by obtaining reference frames provided by AutonomouStuff. GPS coordinates were offset by the current orientation and converted the ground truth to the LiDAR frame. GPS, IMU, and LiDAR reference frames and rotational updates were combined into trajectory vectors. Consecutive point cloud scans may then translated and rotated with derived transformation matrices (Figure \ref{fig:Compiled_PCD}). While not as robust as more sophisticated methods of point cloud aggregation such as NDT or ICP scan matching, over shorter distances using this method proved to be adequate for scoring purposes. Manually defined areas may then be overlayed unto the point cloud (Figure \ref{fig:Combined_Classified_PCD}). Six degree arcs from the first three channels of scanning LiDAR data that lay directly in front of the vehicle were extracted and exported to a database if coincident with the manually defined areas.}

	\begin{figure}
		\centering
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Compiled_PCD}
			\caption[C]{}
			\label{fig:Compiled_PCD}
		\end{subfigure}
		\begin{subfigure}{0.45\textwidth}
			\centering
			\includegraphics[width=0.95\linewidth]{figures/Manually_Defined_Area_Train_Dat_Grab_2}
			\caption[]{}
			\label{fig:Classified_Areas_PCD}
		\end{subfigure}
		\caption[Compiled and Classified Point Cloud Data]{Point cloud of scanning LiDAR data using GPS and IMU data to determine point of origin and orientation for the LiDAR was used to create a combined point cloud map (a). Manually defined areas inform where training data lies (b).}
		\label{fig:Combined_Classified_PCD}
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_percentages_big_anotated}
		\caption[Area Percentage Scores]{Five runs of Blackburn Road with the intercepting gravel parking lot were classified. Shown is the results of one run. Percentages of each terrain type per area were calculated. Area $3$ is the gravel parking lot entrance from which training data was extracted.}
		\label{fig:area_percentages}
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.975\linewidth]{figures/side_of_road_nums_4_with_filts_2}
		\caption[Post-Adjusted]{}
		\label{fig:prepostadjust}
		\caption[Area Scores]{Per-area scores were calculated, indicating algorithm efficiency of unmarked road detection. }
	\end{figure}

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.9\linewidth]{figures/rm_db_4_toc}
		\caption[Projected Guess vs Truth]{Gravel and asphalt surfaces were manually guessed using a rule set and compared to actual road-surface areas. }
		\label{fig:rm_db_4_toc}
	\end{figure}

@@

%%%%% Conclusions %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%% Acknowledgments %%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgments}
Place any acknowledgments here.

%%%  REFERENCES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% Put your references into your .bib file in the usual way. Run latex once, bibtex once, then latex twice.
%% The asmeconf.bst style allows @inproceedings and @proceedings to include: 
%%		venue = {Location of Conference}, 
%%		eventdate = {Month, days},

\nocite{*}%% <=== Delete this line unless you want to typeset the entire contents of your .bib file!

\bibliographystyle{asmeconf}  %% .bst file following ASME conference format. Do not change.
\bibliography{resources.bib}

%%%  APPENDICES  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\end{document}
